---
title: "Median Normalization with SomaScan Data"
author: "SomaScan Bioinformatics Team, SomaLogic, Inc."
description: >
  A guide to median normalization of SomaScan data, including when to use it,
  how to apply it, and quality control considerations.
output:
  rmarkdown::html_vignette:
    fig_caption: yes
vignette: >
  %\VignetteIndexEntry{Median Normalization with SomaScan Data}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  chunk_output_type: console
  markdown: 
    wrap: 72
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "figures/median-normalization-",
  fig.width = 7,
  fig.height = 5
)
```

## What is Median Normalization?

Median normalization is a scale-based normalization technique commonly used
to correct for assay-derived technical variation in SomaScan data. This method
applies sample-specific linear scaling to expression measurements, helping to
improve assay precision and reduce technical variation that can mask true
biological signal.

Median normalization corrects for global intensity shifts by scaling each sample
so that the center of the within-sample analyte distribution aligns to a defined
reference. This approach preserves relative differences between measurements
within a sample while standardizing the overall signal intensity across samples.

Median normalization is particularly useful for:

- **Combining datasets** from the same experiment that were processed separately
- **Re-normalizing data** to a common reference when merging studies  
- **Normalizing sample groups separately** when biological differences are expected
- **Correcting technical variation** from sources like liquid handling,
  consumables, instrument drift, or environmental conditions

**Note**: Median normalization assumes that the *majority of analytes are not differentially expressed* between comparison groups. Users must validate this assumption by inspecting scale factor distributions for systematic bias before proceeding with downstream analyses.

With this assumption in mind, consider avoiding median normalization when:

- A large proportion of analytes are expected to be differentially expressed
- Comparing very different biological states (e.g., healthy vs. severe disease)
- Small sample sizes where medians may be unstable (n < 10 per group)

------------------------------------------------------------------------

## Load Libraries

```{r load-lib, message = FALSE}
library(SomaDataIO)
library(dplyr)
library(ggplot2)
library(tidyr)
```

------------------------------------------------------------------------

## Data Preparation

### Using Example Data

We'll use the built-in `example_data` adat object included in `SomaDataIO`. 
The `example_data` object contains `r nrow(example_data)` samples and
`r length(getAnalytes(example_data))` analytes. It has already been processed
through the standard SomaScan pipeline — the `ProcessSteps` recorded in its
header are:

```{r load-data-process-steps}
attr(example_data, "Header.Meta")$HEADER$ProcessSteps
```

### Reversing Existing Normalization

Note the processing steps show that the study samples have already undergone
ANML normalization (`anmlSMP` as the final step). We will reverse this to 
create unnormalized data that represents what customers typically receive after
basic processing:

```{r create-unnormalized}
# Represents data after HybNorm and PlateScale, before median normalization
unnormalized_adat <- reverseMedianNormalize(example_data)
```

`reverseMedianNormalize()` removes the median normalization step recorded in
`ProcessSteps` and rescales the RFU values back to their pre-normalized state.
The data dimensions are unchanged — only the RFU values and header metadata
are affected. If your ADAT file has not been normalized yet, this step is not required.

### Sample Information

```{r sample-info}
unnormalized_adat |> count(SampleType)
```

For most analyses we focus on `SampleType == "Sample"` rows. The dataset
also includes demographic variables such as sex (`Sex`) and age (`Age`), which 
will be used later to demonstrate group-specific normalization and biological 
signal preservation.

------------------------------------------------------------------------

## Before You Begin

Confirm your data is suitable for median normalization:

- ADAT has complete metadata and valid analyte columns
- Minimal upstream processing present (typically HybNorm + PlateScale)
- Sufficient sample size for stable medians (n >= 10 per group recommended)
- Most analytes expected to be unchanged across comparison groups

Confirm which reference strategy is most appropriate for your data and planned analyses:

- *Internal reference* (default): for single-study normalization
- *External reference*: when anchoring multiple datasets to one target (see Advanced Options)
- *Grouped normalization* (`by=`): when known biology may shift global intensity (see Advanced Options for validation guidance)

Review scale factor acceptance criteria to assess the effects of normalization.
See the [Pre-Processing SomaScan](pre-processing.html) vignette for more details.

------------------------------------------------------------------------

## Basic Usage

The simplest approach uses an internal reference calculated from study samples:

```{r basic-usage}
normalized_adat <- medianNormalize(unnormalized_adat)
```

By default, `medianNormalize()` computes a per-dilution-group reference median
across all study samples. Scale factors centered near 1.0 with low spread
indicate successful normalization.

```{r show-scale-factor-summary}
scale_factor_cols <- grep("^NormScale_", names(normalized_adat), value = TRUE)
normalized_adat[normalized_adat$SampleType == "Sample", scale_factor_cols] |>
  vapply(FUN = summary, FUN.VALUE = numeric(6)) |>
  t()
```

------------------------------------------------------------------------

## Advanced Options

### External Reference Normalization

When combining datasets from different processing batches, normalize each to a 
common external reference to ensure comparable results. The `reference` 
parameter allows you to specify an external dataset whose median will be used 
as the normalization anchor.

For this example, we simulate two separate study batches by randomly splitting 
the original data into two subsets:

```{r multi-study-split}
# Simulate two study batches
set.seed(123)
study1_indices <- sample(which(unnormalized_adat$SampleType == "Sample"), 
                        size = floor(sum(unnormalized_adat$SampleType == "Sample") / 2))

study1_data <- unnormalized_adat[study1_indices, ]
study2_data <- unnormalized_adat[-study1_indices, ]
```

To ensure both studies are on the same scale, we use the full original 
unnormalized dataset as the common external reference, which is passed to 
`medianNormalize()` via the `reference` parameter:

```{r multi-study-normalize}
# Use the full unnormalized dataset as the common external reference
reference_adat <- unnormalized_adat

# Normalize both studies to the same external reference
study1_norm <- medianNormalize(study1_data, reference = reference_adat)
study2_norm <- medianNormalize(study2_data, reference = reference_adat)

# Compare scale factors between studies
study_comparison <- data.frame(
  Study = c(rep("Study 1", nrow(study1_norm)),
            rep("Study 2", nrow(study2_norm))),
  ScaleFactor = c(study1_norm$NormScale_0_5,
                  study2_norm$NormScale_0_5)
)

study_comparison |>
  group_by(Study) |>
  summarise(
    n       = n(),
    Mean_SF = round(mean(ScaleFactor, na.rm = TRUE), 3),
    SD_SF   = round(sd(ScaleFactor, na.rm = TRUE), 3),
    .groups = "drop"
  ) |>
  knitr::kable(caption = "Scale factor statistics by study (normalized to common reference)")
```

```{r multi-study-combine}
# Combine the normalized studies for joint analysis
combined_data <- rbind(study1_norm, study2_norm)
```

Using a common external reference ensures both studies are normalized to the
same scale, making the data directly comparable when combined for analysis.
With an external reference, scale factors reflect how each sample compares
to the provided reference rather than an internal median.

### Grouped Normalization

When biological groups are expected to differ systematically (e.g., disease vs. 
control, treatment vs. placebo), use the `by` parameter to normalize within 
groups. This preserves known biological differences that would otherwise be 
absorbed into a global reference median.

#### When to Use Grouped Normalization

Before using grouped normalization, validate whether your groups differ 
substantially. Examine fold changes between your comparison groups on 
*unnormalized* data. 

**Decision criteria:**

- **< 30% of analytes with >1.5-fold change**: Default pooled normalization is appropriate
- **30-50% of analytes with >1.5-fold change**: Consider grouped normalization
- **> 50% of analytes with >1.5-fold change**: Grouped normalization strongly recommended

For demonstration purposes, we will use the built-in 
binary categorical grouping variable `Sex` that is in the example data.

<details>
<summary>Show code</summary>
```{r assumption-validation-1, echo = TRUE, eval = FALSE}
# Get analytes from unnormalized data
analytes <- getAnalytes(unnormalized_adat)

# Check fold changes using Sex as an example grouping variable
# IMPORTANT: Use unnormalized data for this check
if("Sex" %in% names(unnormalized_adat)) {
  study_data <- unnormalized_adat |> filter(SampleType == "Sample")
  
  # Calculate fold changes between sexes for all analytes
  fold_changes <- sapply(analytes, function(analyte) {
    male_vals <- study_data[[analyte]][study_data$Sex == "M"]
    female_vals <- study_data[[analyte]][study_data$Sex == "F"]
    
    if(length(male_vals) > 0 && length(female_vals) > 0) {
      male_median <- median(male_vals, na.rm = TRUE)
      female_median <- median(female_vals, na.rm = TRUE)
      return(male_median / female_median)
    } else {
      return(NA)
    }
  })
  
  fold_changes <- fold_changes[!is.na(fold_changes)]
  
  if(length(fold_changes) > 0) {
    large_fc_prop <- mean(abs(log2(fold_changes)) > 0.585, na.rm = TRUE)
    
    fc_df <- data.frame(Log2_FC = log2(fold_changes))
    
    ggplot(fc_df, aes(x = Log2_FC)) +
      geom_histogram(bins = 30, alpha = 0.7, fill = "steelblue") +
      geom_vline(xintercept = 0, color = "red", linetype = "dashed", linewidth = 1) +
      geom_vline(xintercept = c(-0.585, 0.585), color = "orange",
                 linetype = "dashed", linewidth = 1) +
      labs(title = "Distribution of Log2 Fold Changes (Male/Female)",
           subtitle = "Red: no change, Orange: ±1.5-fold threshold",
           x = "Log2 Fold Change",
           y = "Count") +
      xlim(-3, 3) +
      theme_minimal()
  }
}
```
</details>

```{r assumption-validation-2, echo = FALSE, warning = FALSE}
# Get analytes from unnormalized data
analytes <- getAnalytes(unnormalized_adat)

# Check fold changes using Sex as an example grouping variable
# IMPORTANT: Use unnormalized data for this check
if("Sex" %in% names(unnormalized_adat)) {
  study_data <- unnormalized_adat |> filter(SampleType == "Sample")
  
  # Calculate fold changes between sexes for all analytes
  fold_changes <- sapply(analytes, function(analyte) {
    male_vals <- study_data[[analyte]][study_data$Sex == "M"]
    female_vals <- study_data[[analyte]][study_data$Sex == "F"]
    
    if(length(male_vals) > 0 && length(female_vals) > 0) {
      male_median <- median(male_vals, na.rm = TRUE)
      female_median <- median(female_vals, na.rm = TRUE)
      return(male_median / female_median)
    } else {
      return(NA)
    }
  })
  
  fold_changes <- fold_changes[!is.na(fold_changes)]
  
  if(length(fold_changes) > 0) {
    large_fc_prop <- mean(abs(log2(fold_changes)) > 0.585, na.rm = TRUE)
    
    fc_df <- data.frame(Log2_FC = log2(fold_changes))
    
    ggplot(fc_df, aes(x = Log2_FC)) +
      geom_histogram(bins = 30, alpha = 0.7, fill = "steelblue") +
      geom_vline(xintercept = 0, color = "red", linetype = "dashed", linewidth = 1) +
      geom_vline(xintercept = c(-0.585, 0.585), color = "orange",
                 linetype = "dashed", linewidth = 1) +
      labs(title = "Distribution of Log2 Fold Changes (Male/Female)",
           subtitle = "Red: no change, Orange: ±1.5-fold threshold",
           x = "Log2 Fold Change",
           y = "Count") +
      xlim(-3, 3) +
      theme_minimal()
  }
}
```

Among the `r length(fold_changes)` analytes evaluated,
`r round(large_fc_prop * 100, 1)`% show a male/female ratio greater than
1.5-fold, well below concerning thresholds. The distribution centered near
zero confirms that grouped normalization by `Sex` is _not required_ for this
dataset.

However, `Sex` can be a useful demonstration variable - we can show how the 
feature works in a scenario where either approach is acceptable. In real
applications requiring grouped normalization (e.g., disease vs. control with 
widespread protein changes), the decision would be clear from the fold change 
distribution.

#### Applying Grouped Normalization

To normalize by Sex:

```{r custom-grouping}
grouped_norm_adat <- medianNormalize(unnormalized_adat, by = "Sex")
```

Using `by = "Sex"` computes separate reference medians for each sex group
and scales samples within their own group. This preserves known sex-linked
differences (e.g., sex-hormone proteins) that would otherwise be absorbed into
a global reference median.

```{r compare-grouped-sf}
data.frame(
  SampleId   = normalized_adat$SampleId[1:8],
  Sex        = normalized_adat$Sex[1:8],
  Default_SF = normalized_adat$NormScale_0_5[1:8],
  Grouped_SF = grouped_norm_adat$NormScale_0_5[1:8]
)
```

To demonstrate how grouped normalization works, we can examine 
Prostate-specific antigen (PSA), which shows strong biological differences 
between males and females:

<details>
<summary>Show code</summary>
```{r grouped-normalization-validation-1, echo = TRUE, eval = FALSE}
# PSA is a well-known sex-associated protein
psa_analyte <- "seq.8468.19"  # Prostate-specific antigen

# Calculate the effect of Sex on PSA across normalization approaches
calc_sex_effect <- function(adat, label) {
  study_data <- adat |> filter(SampleType == "Sample")
  male_values <- study_data[[psa_analyte]][study_data$Sex == "M"]
  female_values <- study_data[[psa_analyte]][study_data$Sex == "F"]
  
  # Calculate fold change
  fold_change <- mean(male_values, na.rm = TRUE) / mean(female_values, na.rm = TRUE)
  
  # Perform t-test
  t_result <- t.test(male_values, female_values)
  
  data.frame(
    State = label,
    Male_mean = round(mean(male_values, na.rm = TRUE), 0),
    Female_mean = round(mean(female_values, na.rm = TRUE), 0),
    Fold_change = round(fold_change, 2),
    P_value = ifelse(t_result$p.value < 0.001, "<0.001", round(t_result$p.value, 3)),
    stringsAsFactors = FALSE
  )
}

sex_comparison <- rbind(
  calc_sex_effect(unnormalized_adat, "Unnormalized"),
  calc_sex_effect(normalized_adat, "Default (pooled)"),
  calc_sex_effect(grouped_norm_adat, "Grouped (by Sex)")
)

knitr::kable(sex_comparison, 
             caption = "PSA levels by sex across normalization approaches",
             col.names = c("Normalization", "Male Mean", "Female Mean", "Fold Change", "p-value"))
```
</details>

```{r grouped-normalization-validation-2, echo = FALSE}
# PSA is a well-known sex-associated protein
psa_analyte <- "seq.8468.19"  # Prostate-specific antigen

# Calculate the effect of Sex on PSA across normalization approaches
calc_sex_effect <- function(adat, label) {
  study_data <- adat |> filter(SampleType == "Sample")
  male_values <- study_data[[psa_analyte]][study_data$Sex == "M"]
  female_values <- study_data[[psa_analyte]][study_data$Sex == "F"]
  
  # Calculate fold change
  fold_change <- mean(male_values, na.rm = TRUE) / mean(female_values, na.rm = TRUE)
  
  # Perform t-test
  t_result <- t.test(male_values, female_values)
  
  data.frame(
    State = label,
    Male_mean = round(mean(male_values, na.rm = TRUE), 0),
    Female_mean = round(mean(female_values, na.rm = TRUE), 0),
    Fold_change = round(fold_change, 2),
    P_value = ifelse(t_result$p.value < 0.001, "<0.001", round(t_result$p.value, 3)),
    stringsAsFactors = FALSE
  )
}

sex_comparison <- rbind(
  calc_sex_effect(unnormalized_adat, "Unnormalized"),
  calc_sex_effect(normalized_adat, "Default (pooled)"),
  calc_sex_effect(grouped_norm_adat, "Grouped (by Sex)")
)

knitr::kable(sex_comparison, 
             caption = "PSA levels by sex across normalization approaches",
             col.names = c("Normalization", "Male Mean", "Female Mean", "Fold Change", "p-value"))
```

Males have ~7-fold higher PSA levels than females, reflecting the biological
reality that PSA is primarily produced by the prostate gland. This example 
illustrates how grouped normalization (fold change = 
`r sex_comparison$Fold_change[3]`) can preserve biological differences more 
accurately than default pooled normalization (fold change = 
`r sex_comparison$Fold_change[2]`), which slightly inflates the effect. In 
scenarios where grouped normalization is truly required, these differences 
would be more pronounced.

------------------------------------------------------------------------

## Quality Assessment

Median normalization should reduce technical variability while preserving 
biological signal. Successful normalization will show:

- **Reduced coefficient of variation (CV)** across analytes (lower CV indicates better precision)
- **Scale factors clustered near 1.0** with minimal spread (indicating consistent normalization)
- **Preserved biological patterns** in sample relationships (e.g., expected groupings remain visible)

Review the metrics below to confirm normalization improved data quality without over-correction.

### Population-Level Effects

<details>
<summary>Show code</summary>
```{r population-comparison-1, echo = TRUE, eval = FALSE}
# Calculate population-level statistics for each state
calc_population_stats <- function(adat, label) {
  study_data <- adat |> filter(SampleType == "Sample")
  analyte_data <- study_data[, getAnalytes(adat)]
  
  data.frame(
    State = label,
    n_samples = nrow(study_data),
    mean_RFU = round(mean(unlist(analyte_data), na.rm = TRUE), 0),
    median_RFU = round(median(unlist(analyte_data), na.rm = TRUE), 0),
    cv_RFU = round(sd(unlist(analyte_data), na.rm = TRUE) / 
                   mean(unlist(analyte_data), na.rm = TRUE) * 100, 2),
    stringsAsFactors = FALSE
  )
}

comparison_stats <- rbind(
  calc_population_stats(unnormalized_adat, "Unnormalized"),
  calc_population_stats(normalized_adat, "Median Normalized")
)

knitr::kable(comparison_stats, 
             caption = "Population-level RFU statistics across normalization states")
```
</details>

```{r population-comparison-2, echo = FALSE}
# Calculate population-level statistics for each state
calc_population_stats <- function(adat, label) {
  study_data <- adat |> filter(SampleType == "Sample")
  analyte_data <- study_data[, getAnalytes(adat)]
  
  data.frame(
    State = label,
    n_samples = nrow(study_data),
    mean_RFU = round(mean(unlist(analyte_data), na.rm = TRUE), 0),
    median_RFU = round(median(unlist(analyte_data), na.rm = TRUE), 0),
    cv_RFU = round(sd(unlist(analyte_data), na.rm = TRUE) / 
                   mean(unlist(analyte_data), na.rm = TRUE) * 100, 2),
    stringsAsFactors = FALSE
  )
}

comparison_stats <- rbind(
  calc_population_stats(unnormalized_adat, "Unnormalized"),
  calc_population_stats(normalized_adat, "Median Normalized")
)

knitr::kable(comparison_stats, 
             caption = "Population-level RFU statistics across normalization states")
```

Median normalization reduces the population-level CV compared to unnormalized 
data, indicating improved technical consistency. Note that mean and median RFU 
values remain similar, showing that signal intensity is preserved.

### Scale Factor Distribution

<details>
<summary>Show code</summary>
```{r scale-factor-plot-1, echo = TRUE, eval = FALSE}
sf_cols <- grep("^NormScale_", names(normalized_adat), value = TRUE)

study_sf <- normalized_adat |>
  filter(SampleType == "Sample") |>
  select(SampleId, all_of(sf_cols)) |>
  pivot_longer(all_of(sf_cols), names_to = "Dilution", values_to = "ScaleFactor") |>
  mutate(Dilution = gsub("NormScale_", "", Dilution))

med_sf <- study_sf |>
  group_by(Dilution) |>
  summarise(med = median(ScaleFactor, na.rm = TRUE), .groups = "drop")

ggplot(study_sf, aes(x = ScaleFactor)) +
  geom_histogram(bins = 30, alpha = 0.7, fill = "steelblue") +
  geom_vline(xintercept = 1, color = "red", linetype = "dashed") +
  geom_text(data = med_sf, aes(x = Inf, y = Inf, 
                                label = paste0("Median: ", round(med, 3))),
            hjust = 1.1, vjust = 1.5, size = 3, color = "black") +
  facet_wrap(~Dilution, scales = "free_x") +
  labs(title = "Normalization Scale Factors by Dilution Group",
       subtitle = "Red line: 1.0 reference",
       x = "Scale Factor", y = "Count") +
  theme_minimal()
```
</details>

```{r scale-factor-plot-2, fig.width=8, fig.height=5, echo = FALSE}
sf_cols <- grep("^NormScale_", names(normalized_adat), value = TRUE)

study_sf <- normalized_adat |>
  filter(SampleType == "Sample") |>
  select(SampleId, all_of(sf_cols)) |>
  pivot_longer(all_of(sf_cols), names_to = "Dilution", values_to = "ScaleFactor") |>
  mutate(Dilution = gsub("NormScale_", "", Dilution))

med_sf <- study_sf |>
  group_by(Dilution) |>
  summarise(med = median(ScaleFactor, na.rm = TRUE), .groups = "drop")

ggplot(study_sf, aes(x = ScaleFactor)) +
  geom_histogram(bins = 30, alpha = 0.7, fill = "steelblue") +
  geom_vline(xintercept = 1, color = "red", linetype = "dashed") +
  geom_text(data = med_sf, aes(x = Inf, y = Inf, 
                                label = paste0("Median: ", round(med, 3))),
            hjust = 1.1, vjust = 1.5, size = 3, color = "black") +
  facet_wrap(~Dilution, scales = "free_x") +
  labs(title = "Normalization Scale Factors by Dilution Group",
       subtitle = "Red line: 1.0 reference",
       x = "Scale Factor", y = "Count") +
  theme_minimal()
```

Scale factors should cluster near 1.0. Samples with factors below 0.5 or
above 2.0 warrant investigation. See the
[Pre-Processing SomaScan](pre-processing.html) vignette for detailed
QC criteria and the `RowCheck` column discussion.

If scale factors show systematic patterns (e.g., consistently shifted from 1.0
across all samples or specific groups), this may indicate:

- **Median slightly above/below 1.0 but tightly distributed**: Normal variation, 
  no concern if all samples cluster together
- **Bimodal distribution**: May indicate batch effects or sample collection 
  differences - investigate whether distinct groups correspond to known technical 
  or biological factors
- **Consistent shift for specific dilutions**: Could reflect dilution-specific 
  technical issues during sample processing
- **Wide spread (many samples < 0.5 or > 2.0)**: Suggests poor quality samples 
  or reference mismatch - review sample collection protocols and consider 
  removing outlier samples

### Precision Improvement

<details>
<summary>Show code</summary>
```{r cv-reduction-plot-1, echo = TRUE, eval = FALSE}
# Get analytes
analytes <- getAnalytes(unnormalized_adat)

# Calculate CV for all analytes before and after normalization
study_unnorm <- unnormalized_adat |> filter(SampleType == "Sample")
study_norm <- normalized_adat |> filter(SampleType == "Sample")

cv_unnorm <- sapply(analytes, function(x) {
  vals <- study_unnorm[[x]]
  sd(vals, na.rm = TRUE) / mean(vals, na.rm = TRUE) * 100
})

cv_norm <- sapply(analytes, function(x) {
  vals <- study_norm[[x]]
  sd(vals, na.rm = TRUE) / mean(vals, na.rm = TRUE) * 100
})

# Create CV comparison data
cv_data <- data.frame(
  CV = c(cv_unnorm, cv_norm),
  State = rep(c("Unnormalized", "Median Normalized"), each = length(cv_unnorm)),
  stringsAsFactors = FALSE
)
cv_data$State <- factor(cv_data$State, levels = c("Unnormalized", "Median Normalized"))

# Plot CV distributions
ggplot(cv_data, aes(x = CV, fill = State)) +
  geom_histogram(alpha = 0.7, bins = 50, position = "identity") +
  scale_fill_manual(values = c("Unnormalized" = "#E69F00", "Median Normalized" = "#56B4E9")) +
  labs(title = "Distribution of Analyte Coefficients of Variation",
       subtitle = "Lower CV indicates better precision",
       x = "Coefficient of Variation (%)",
       y = "Count") +
  theme_minimal()

# Summary statistics
cv_summary <- cv_data |>
  group_by(State) |>
  summarise(
    Mean_CV = round(mean(CV, na.rm = TRUE), 2),
    Median_CV = round(median(CV, na.rm = TRUE), 2),
    SD_CV = round(sd(CV, na.rm = TRUE), 2),
    .groups = "drop"
  )

knitr::kable(cv_summary, 
             caption = "Coefficient of Variation summary statistics")
```
</details>

```{r cv-reduction-plot-2, fig.width=8, fig.height=6, echo = FALSE}
# Get analytes
analytes <- getAnalytes(unnormalized_adat)

# Calculate CV for all analytes before and after normalization
study_unnorm <- unnormalized_adat |> filter(SampleType == "Sample")
study_norm <- normalized_adat |> filter(SampleType == "Sample")

cv_unnorm <- sapply(analytes, function(x) {
  vals <- study_unnorm[[x]]
  sd(vals, na.rm = TRUE) / mean(vals, na.rm = TRUE) * 100
})

cv_norm <- sapply(analytes, function(x) {
  vals <- study_norm[[x]]
  sd(vals, na.rm = TRUE) / mean(vals, na.rm = TRUE) * 100
})

# Create CV comparison data
cv_data <- data.frame(
  CV = c(cv_unnorm, cv_norm),
  State = rep(c("Unnormalized", "Median Normalized"), each = length(cv_unnorm)),
  stringsAsFactors = FALSE
)
cv_data$State <- factor(cv_data$State, levels = c("Unnormalized", "Median Normalized"))

# Plot CV distributions
ggplot(cv_data, aes(x = CV, fill = State)) +
  geom_histogram(alpha = 0.7, bins = 50, position = "identity") +
  scale_fill_manual(values = c("Unnormalized" = "#E69F00", "Median Normalized" = "#56B4E9")) +
  labs(title = "Distribution of Analyte Coefficients of Variation",
       subtitle = "Lower CV indicates better precision",
       x = "Coefficient of Variation (%)",
       y = "Count") +
  theme_minimal()

# Summary statistics
cv_summary <- cv_data |>
  group_by(State) |>
  summarise(
    Mean_CV = round(mean(CV, na.rm = TRUE), 2),
    Median_CV = round(median(CV, na.rm = TRUE), 2),
    SD_CV = round(sd(CV, na.rm = TRUE), 2),
    .groups = "drop"
  )

knitr::kable(cv_summary, 
             caption = "Coefficient of Variation summary statistics")
```

The leftward shift in CV distribution indicates improved technical
consistency after normalization. Both mean and median CV values are reduced 
for median normalized data compared to unnormalized, demonstrating the 
effectiveness of the normalization procedure.

### Biological Signal Preservation

An important quality check is confirming that normalization removes technical
variation without distorting biological relationships. We can verify this by
examining analytes known to associate with age. The example data includes an 
`Age` variable that we'll use to test whether age-associated correlations are 
preserved after normalization:

<details>
<summary>Show code</summary>
```{r age-correlation-1, echo = TRUE, eval = FALSE}
# Select analytes with known age associations
age_analytes <- c(
  "seq.3045.72",   # Pleiotrophin
  "seq.4374.45",   # Growth/differentiation factor 15
  "seq.15640.54",  # Transgelin
  "seq.6392.7",    # WISP2
  "seq.4496.60"    # Macrophage metalloelastase
)

# Calculate correlations with age before and after normalization
age_cors_unnorm <- sapply(age_analytes, function(apt) {
  cor(study_unnorm[[apt]], study_unnorm$Age, use = "complete.obs")
})

age_cors_norm <- sapply(age_analytes, function(apt) {
  cor(study_norm[[apt]], study_norm$Age, use = "complete.obs")
})

# Test if correlations differ significantly using Fisher's z-transformation
n <- sum(complete.cases(study_unnorm$Age))
z_diff <- sapply(seq_along(age_analytes), function(i) {
  # Fisher's z-transformation
  z1 <- 0.5 * log((1 + age_cors_unnorm[i]) / (1 - age_cors_unnorm[i]))
  z2 <- 0.5 * log((1 + age_cors_norm[i]) / (1 - age_cors_norm[i]))
  # Test statistic
  z_diff <- (z1 - z2) / sqrt(2 / (n - 3))
  # Two-tailed p-value
  p_val <- 2 * (1 - pnorm(abs(z_diff)))
  return(p_val)
})

# Get target names
target_names <- getAnalyteInfo(unnormalized_adat) |>
  filter(AptName %in% age_analytes) |>
  select(AptName, Target = TargetFullName)

# Create comparison table
age_cor_comparison <- data.frame(
  Target = target_names$Target[match(age_analytes, target_names$AptName)],
  Cor_Unnormalized = round(age_cors_unnorm, 3),
  Cor_Normalized = round(age_cors_norm, 3),
  Difference = round(age_cors_norm - age_cors_unnorm, 3),
  P_value = ifelse(z_diff < 0.001, "<0.001", round(z_diff, 3))
)

knitr::kable(age_cor_comparison,
             caption = "Age correlations before and after median normalization",
             col.names = c("Target", "Unnormalized", "Median Normalized", "Difference", "p-value"))

# Visualize age association for top analyte
top_analyte <- age_analytes[1]
top_target <- target_names$Target[target_names$AptName == top_analyte]

# Create long-format data for plotting
plot_data <- data.frame(
  Age = rep(study_unnorm$Age, 2),
  RFU = c(study_unnorm[[top_analyte]], study_norm[[top_analyte]]),
  State = rep(c("Unnormalized", "Median Normalized"), each = nrow(study_unnorm))
)
plot_data$State <- factor(plot_data$State, levels = c("Unnormalized", "Median Normalized"))

ggplot(plot_data, aes(x = Age, y = RFU, color = State)) +
  geom_point(alpha = 0.5, size = 2) +
  geom_smooth(method = "lm", se = TRUE, linewidth = 1) +
  scale_color_manual(values = c("Unnormalized" = "#E69F00", "Median Normalized" = "#56B4E9")) +
  labs(title = paste("Age Association:", top_target),
       subtitle = paste0("Unnormalized r = ", round(age_cors_unnorm[1], 3), 
                        ", Normalized r = ", round(age_cors_norm[1], 3)),
       x = "Age (years)",
       y = "RFU",
       color = "Normalization") +
  theme_minimal() +
  theme(legend.position = "bottom")
```
</details>

```{r age-correlation-2, echo = FALSE, fig.width=8, fig.height=6}
# Select analytes with known age associations
age_analytes <- c(
  "seq.3045.72",   # Pleiotrophin
  "seq.4374.45",   # Growth/differentiation factor 15
  "seq.15640.54",  # Transgelin
  "seq.6392.7",    # WISP2
  "seq.4496.60"    # Macrophage metalloelastase
)

# Calculate correlations with age before and after normalization
age_cors_unnorm <- sapply(age_analytes, function(apt) {
  cor(study_unnorm[[apt]], study_unnorm$Age, use = "complete.obs")
})

age_cors_norm <- sapply(age_analytes, function(apt) {
  cor(study_norm[[apt]], study_norm$Age, use = "complete.obs")
})

# Test if correlations differ significantly using Fisher's z-transformation
n <- sum(complete.cases(study_unnorm$Age))
z_diff <- sapply(seq_along(age_analytes), function(i) {
  # Fisher's z-transformation
  z1 <- 0.5 * log((1 + age_cors_unnorm[i]) / (1 - age_cors_unnorm[i]))
  z2 <- 0.5 * log((1 + age_cors_norm[i]) / (1 - age_cors_norm[i]))
  # Test statistic
  z_diff <- (z1 - z2) / sqrt(2 / (n - 3))
  # Two-tailed p-value
  p_val <- 2 * (1 - pnorm(abs(z_diff)))
  return(p_val)
})

# Get target names
target_names <- getAnalyteInfo(unnormalized_adat) |>
  filter(AptName %in% age_analytes) |>
  select(AptName, Target = TargetFullName)

# Create comparison table
age_cor_comparison <- data.frame(
  Target = target_names$Target[match(age_analytes, target_names$AptName)],
  Cor_Unnormalized = round(age_cors_unnorm, 3),
  Cor_Normalized = round(age_cors_norm, 3),
  Difference = round(age_cors_norm - age_cors_unnorm, 3),
  P_value = ifelse(z_diff < 0.001, "<0.001", round(z_diff, 3))
)

knitr::kable(age_cor_comparison,
             caption = "Age correlations before and after median normalization",
             col.names = c("Target", "Unnormalized", "Median Normalized", "Difference", "p-value"))
```

The age correlations remain strong and statistically unchanged after 
normalization (all p-values > 0.05), demonstrating that the biological 
relationships are preserved. While correlation coefficients may shift slightly, 
Fisher's z-test confirms these differences are not statistically significant. 
To visualize this relationship more clearly, we examine a scatter plot for the 
analyte with the strongest correlation, Pleiotrophin:

```{r age-scatterplot, echo = FALSE, fig.width=8, fig.height=6, message = FALSE}
# Visualize age association for top analyte
top_analyte <- age_analytes[1]
top_target <- target_names$Target[target_names$AptName == top_analyte]

# Create long-format data for plotting
plot_data <- data.frame(
  Age = rep(study_unnorm$Age, 2),
  RFU = c(study_unnorm[[top_analyte]], study_norm[[top_analyte]]),
  State = rep(c("Unnormalized", "Median Normalized"), each = nrow(study_unnorm))
)
plot_data$State <- factor(plot_data$State, levels = c("Unnormalized", "Median Normalized"))

ggplot(plot_data, aes(x = Age, y = RFU, color = State)) +
  geom_point(alpha = 0.5, size = 2) +
  geom_smooth(method = "lm", se = TRUE, linewidth = 1) +
  scale_color_manual(values = c("Unnormalized" = "#E69F00", "Median Normalized" = "#56B4E9")) +
  labs(title = paste("Age Association:", top_target),
       subtitle = paste0("Unnormalized r = ", round(age_cors_unnorm[1], 3), 
                        ", Normalized r = ", round(age_cors_norm[1], 3)),
       x = "Age (years)",
       y = "RFU",
       color = "Normalization") +
  theme_minimal() +
  theme(legend.position = "bottom")
```

The scatter plot shows that the biological relationship is preserved - both 
regression lines have similar slopes, confirming that median normalization 
successfully removes technical variation without obscuring the age-associated 
signal.

### Comparison to ANML Normalization

The example data was originally normalized using ANML (Adaptive 
Normalization via Maximum-Likelihood), the standard normalization for
most SomaScan deliveries. How does median normalization compare?

<details>
<summary>Show code</summary>
```{r anml-comparison-1, echo = TRUE, eval = FALSE}
# Calculate CVs for ANML-normalized data
study_anml <- example_data |> filter(SampleType == "Sample")

cv_anml <- sapply(analytes, function(x) {
  vals <- study_anml[[x]]
  sd(vals, na.rm = TRUE) / mean(vals, na.rm = TRUE) * 100
})

# Create 3-way comparison data
cv_comp_data <- data.frame(
  CV = c(cv_unnorm, cv_anml, cv_norm),
  Method = rep(c("Unnormalized", "ANML", "Median"), each = length(analytes)),
  stringsAsFactors = FALSE
)
cv_comp_data$Method <- factor(cv_comp_data$Method, 
                               levels = c("Unnormalized", "ANML", "Median"))

# Plot 3-way comparison
ggplot(cv_comp_data, aes(x = CV, fill = Method)) +
  geom_histogram(alpha = 0.6, bins = 50, position = "identity") +
  scale_fill_manual(values = c("Unnormalized" = "#E69F00", 
                                "ANML" = "#56B4E9", 
                                "Median" = "#009E73")) +
  labs(title = "Comparison of Normalization Methods",
       subtitle = "Distribution of Analyte Coefficients of Variation",
       x = "Coefficient of Variation (%)",
       y = "Count") +
  theme_minimal() +
  xlim(0, 200)

# Summary table
method_summary <- cv_comp_data |>
  group_by(Method) |>
  summarise(
    Mean_CV = round(mean(CV, na.rm = TRUE), 2),
    Median_CV = round(median(CV, na.rm = TRUE), 2),
    Pct_Below_50 = round(100 * sum(CV < 50, na.rm = TRUE) / n(), 1),
    .groups = "drop"
  )

knitr::kable(method_summary,
             caption = "Normalization method comparison",
             col.names = c("Method", "Mean CV (%)", "Median CV (%)", "% Analytes CV < 50%"))
```
</details>

```{r anml-comparison-2, fig.width=8, fig.height=6, echo = FALSE, warning = FALSE}
# Calculate CVs for ANML-normalized data
study_anml <- example_data |> filter(SampleType == "Sample")

cv_anml <- sapply(analytes, function(x) {
  vals <- study_anml[[x]]
  sd(vals, na.rm = TRUE) / mean(vals, na.rm = TRUE) * 100
})

# Create 3-way comparison data
cv_comp_data <- data.frame(
  CV = c(cv_unnorm, cv_anml, cv_norm),
  Method = rep(c("Unnormalized", "ANML", "Median"), each = length(analytes)),
  stringsAsFactors = FALSE
)
cv_comp_data$Method <- factor(cv_comp_data$Method, 
                               levels = c("Unnormalized", "ANML", "Median"))

# Plot 3-way comparison
ggplot(cv_comp_data, aes(x = CV, fill = Method)) +
  geom_histogram(alpha = 0.6, bins = 50, position = "identity") +
  scale_fill_manual(values = c("Unnormalized" = "#E69F00", 
                                "ANML" = "#56B4E9", 
                                "Median" = "#009E73")) +
  labs(title = "Comparison of Normalization Methods",
       subtitle = "Distribution of Analyte Coefficients of Variation",
       x = "Coefficient of Variation (%)",
       y = "Count") +
  theme_minimal() +
  xlim(0, 200)

# Summary table
method_summary <- cv_comp_data |>
  group_by(Method) |>
  summarise(
    Mean_CV = round(mean(CV, na.rm = TRUE), 2),
    Median_CV = round(median(CV, na.rm = TRUE), 2),
    Pct_Below_50 = round(100 * sum(CV < 50, na.rm = TRUE) / n(), 1),
    .groups = "drop"
  )

knitr::kable(method_summary,
             caption = "Normalization method comparison",
             col.names = c("Method", "Mean CV (%)", "Median CV (%)", "% Analytes CV < 50%"))
```

Both ANML and median normalization substantially reduce technical
variation compared to unnormalized data. In this dataset, median
normalization achieves slightly better precision 
(`r method_summary$Pct_Below_50[method_summary$Method == "Median"]`% vs 
`r method_summary$Pct_Below_50[method_summary$Method == "ANML"]`% of
analytes with CV < 50%).

**When to choose:**

- **ANML**: Standard for most SomaScan deliveries; uses adaptive reference 
  from large healthy population; recommended when combining data with 
  existing ANML-normalized datasets
- **Median normalization**: Useful for re-normalizing to study-specific 
  reference, combining separately processed batches, or normalizing groups 
  separately when biological differences are expected

For most analyses on standard deliveries, either method is appropriate.

------------------------------------------------------------------------

## Questions

As always, if you have any normalization questions,
we are here to help. Please reach out to us via:

* GitHub [SUPPORT](https://somalogic.github.io/SomaDataIO/SUPPORT.html)
* Global Scientific Engagement Team: <techsupport@somalogic.com>
* General SomaScan inquiries: <support@somalogic.com>
